---
title: "Thesis_GSS"
---

# Research question:

How do physicochemical indicators of soil alter between forests, peatland, grassland in the Netherlands?

Furthermore, what is the relationship between vegetation characteristics that occur in each habitat and the respective physicochemical indicators?  

# Data analysis setup:

## Installing packages

```{r}
install.packages("rstatix")
install.packages("metan")
install.packages("Hmsic")
install.packages("emmeans")
install.packages("ggpubr")
install.packages("cowplot")

```

## Library

```{library(knitr)}
library(dunn.test)
library(corrplot)
library(ggcorrplot)
library(ggplot2)
library(rstatix)
library(readxl)
library(dplyr)
library(psych)
library(vegan)
library(car)
library(emmeans)
library(reshape2)
library(ggpubr)
library(cowplot)
library(gridGraphics)

```

## Loading data

```{r}
soil_health_data<-read_xlsx("Thesis_Soil-health_Raw-data .xlsx")
vegitation_data<-read_xlsx("Thesis_Vegitation_charactersitics.xlsx")
peat_cor_data<-read_xlsx("Soil_peat_data.xlsx")
forest_cor_data<-read_xlsx("Soil_forest_data.xlsx")
grass_cor_data<-read_xlsx("Soil_grass_data.xlsx")
all_cor_data<-read_xlsx("Soil_all_data.xlsx")
```

## Vegetation data preparation

```{r}
vegitation_data<-vegitation_data |> 
  filter(Species != "Litter")

veg_join_data <- vegitation_data |> 
  group_by(Site,Plot, Species) |> 
  summarise(Avg_cover=mean(Coverage)) |>
  mutate(Site = as.factor(Site)) |> 
  pivot_wider(names_from = Species,
              values_from = Avg_cover,
              values_fill = 0) |> 
  mutate(Total = sum(c_across("Anthoxanthum Odoxathum":"Vicia"))) |> 
  pivot_longer(cols ="Anthoxanthum Odoxathum":"Vicia",
               names_to = "Plant_species", 
               values_to = "Abundance") |> 
  filter(Abundance>0) |> 
  mutate(Rel_abundance = Abundance/Total) |> 
  group_by(Site, Plot) |> 
  summarise(q0=sum(Rel_abundance^0)^(1/(1-0)),
            q1=exp(-sum(Rel_abundance*log(Rel_abundance))), 
            q2= sum(Rel_abundance^2)^(1/(1-2)))
```

# Generalized linear model

## Vegetation characteristics

### fitting a GLM to vegetation characteristics

```{r}

levels(veg_join_data$Site)<- c("Peatland", "Forest", "Grassland")
veg_join_data<- veg_join_data |> 
rename(Habitat_type = Site)

q0_glm<-glm(q0~Habitat_type, 
            data = veg_join_data, 
            family = quasipoisson(link = "log"))

veg_join_data$q0_residuals <- residuals(q0_glm)
veg_join_data$q0_fitted <- fitted(q0_glm)

ggplot(veg_join_data, aes(x = q0_fitted,
                 y = q0_residuals)) +   
    geom_point() +    
    geom_hline(yintercept = 0,              
               linetype = "dashed")+ 
    theme_bw() +   
    xlab("Fitted values")+   
    ylab("Residuals")+   
    ggtitle("Residual plot") 

summary(q0_glm)
null_deviance <- summary(q0_glm)$null.deviance
residual_deviance <- summary(q0_glm)$deviance

100*(null_deviance-residual_deviance)/null_deviance

Anova(q0_glm)

summary(emmeans(object = q0_glm, 
                         specs = "Site", 
                         contr = "pairwise"), 
         type="response")

plant_mean<- plant_data |>  group_by(Site) |> 
  summarise(mean_q0 = mean(q0),
            mean_q1= mean(q2),
            mean_q2= mean(q1))
  
#--------------------------------------Shannons Diversity--------------------------------#
q1_glm<-glm(q1~Habitat_type, 
            data = veg_join_data, 
            family = Gamma(link = "log"))

veg_join_data$q1_resdiuals <- residuals(q1_glm)
veg_join_data$q1_fitted <- fitted(q1_glm)

ggplot(veg_join_data, aes(x = q1_fitted,
                 y = q1_residuals)) +   
    geom_point() +    
    geom_hline(yintercept = 0,              
               linetype = "dashed")+ 
    theme_bw() +   
    xlab("Fitted values")+   
    ylab("Residuals")+   
    ggtitle("Residual plot") 

summary(q1_glm)
null_deviance <- summary(q1_glm)$null.deviance
residual_deviance <- summary(q1_glm)$deviance

100*(null_deviance-residual_deviance)/null_deviance

Anova(q1_glm)

summary(emmeans(object = q1_glm, 
                         specs = "Site", 
                         contr = "pairwise"), 
         type="response")


#-----------------------------------------Simpsons diversity ----------------------------#
q2_glm<-glm(q2~Habitat_type, 
            data = veg_join_data, 
            family = Gamma(link = "log"))

veg_join_data$q2_resdiuals <- residuals(q2_glm)
veg_join_data$q2_fitted <- fitted(q2_glm)

ggplot(veg_join_data, aes(x = q2_fitted,
                 y = q2_residuals)) +   
    geom_point() +    
    geom_hline(yintercept = 0,              
               linetype = "dashed")+ 
    theme_bw() +   
    xlab("Fitted values")+   
    ylab("Residuals")+   
    ggtitle("Residual plot") 

summary(q2_glm)
null_deviance <- summary(q2_glm)$null.deviance
residual_deviance <- summary(q2_glm)$deviance

100*(null_deviance-residual_deviance)/null_deviance

Anova(q2_glm)

summary(emmeans(object = q2_glm, 
                         specs = "Site", 
                         contr = "pairwise"), 
         type="response")
  
```

### Plotting vegetation characteristics data

```{r}

levels(veg_data_hill$Site)<- c("Peatland", "Forest", "Grassland")

veg_post_hoc <- data.frame(
  Site = rep(c("Peatland", "Forest", "Grassland"), 3),
  q = rep(c(0, 1, 2), each = 3),
  y = c(rep(12, 3),
        rep(8, 3),
        rep(6, 3)),
  Label = c("a", "b", "a",
            "a", "b", "a",
            "ab", "b", "a"))

veg_plot<-ggplot(veg_data_hill,
       aes(x=Habitat_type, 
           y=diversity, color=Habitat_type))+
  scale_color_manual(values = c("#365D8DFF","#20A486FF","red"))+
  geom_jitter(shape = 15, 
              alpha = 0.75,
              height = 0, 
              width = 0.2)+
  stat_summary(fun.data = "mean_cl_boot")+
  theme_bw()+
  xlab("")+
  ylab("Effective number of species")+
  facet_wrap(~q, ncol=3)+
  theme(axis.text.x=element_text(angle = 45,
                                 vjust = 1,
                                 hjust = 1))+
   theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))+
  geom_text(data=veg_post_hoc,
            aes(y=y,
                x=Site,
                label=Label),
            color="black")+
  ggtitle("Vegetation diversity across three habitat types for each hill number")+
  theme(plot.title = element_text(size = 10))

ggsave(filename = "veg_plot.png",
       plot = veg_plot,
       dpi = 1000,
       width = 20,
       height=12,
       units="cm")


  
```

## Chemical characteristics

### fitting a GLM to chemical characteristics

```{r}

#-------------------------------------------pH----------------------------------------------#
soil_health_fixed <- soil_health_data |> 
  rename(Habitat_type = Site) |> 
  rename(Soil_depth = `Soil depth`)


soil_health_fixed$Habitat_type<-as.factor(soil_health_fixed$Habitat_type)
soil_health_fixed$Soil_depth<-as.factor(soil_health_fixed$Soil_depth)

levels(soil_health_fixed$Habitat_type)<- c("Peatland", "Forest", "Grassland")



pH_glm<-glm(pH~Habitat_type*Soil_depth, 
            data = soil_health_fixed, 
            family = Gamma(link = "log"))

pH_EMM<-emmeans(pH_glm,~Site*Soil_depth)

summary(pH_EMM)
pairs(pH_EMM)
summary(pH_glm)


null_deviance <- summary(pH_glm)$null.deviance
residual_deviance <- summary(pH_glm)$deviance

100*(null_deviance-residual_deviance)/null_deviance

Anova(pH_glm)


#-------------------------------------------EC----------------------------------------------#
EC_glm<-glm(EC~Habitat_type*Soil_depth, 
            data = soil_health_fixed, 
            family = Gamma(link = "log"))

EC_EMM<-emmeans(EC_glm,~Habitat_type*Soil_depth)

summary(EC_glm)
summary(EC_EMM)
pairs(EC_EMM)



null_deviance <- summary(EC_glm)$null.deviance
residual_deviance <- summary(EC_glm)$deviance

100*(null_deviance-residual_deviance)/null_deviance

Anova(EC_glm)


```

### Plotting chemical characteristics from GLM

```{r}

##-----------Plotting the relationship of soil EC, soil depth and habitat type----------##
pH_post_hoc <- data.frame(
  Habitat_type = rep(c("Peatland", "Forest", "Grassland"), 2),
  Soil_depth = rep(c("0-15 cm","15-30 cm" ), each = 3),
  y = c(rep(8, 3),
        rep(8, 3)),
  Label = c("a,A", "b,A", "a,A",
            "a,A","b,A","a,A"))

ggplot(soil_health_fixed, 
       aes(x = Habitat_type, y= pH, fill= Habitat_type))+ 
  xlab("")+
  ylab("Soil pH")+
  theme_bw()+
  facet_wrap(~Soil_depth)+
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))+
  geom_boxplot()+
  scale_fill_manual(values = c("#365D8DFF","#20A486FF","red"))+
  geom_text(data=pH_post_hoc,
            aes(y=y,
                x=Habitat_type,
                label=Label),
            color="black")+
  ggtitle("Soil pH across soil depths and habitat type across Noorder park")+
    theme(plot.title = element_text(size = 10))

ggsave(filename = "pH_plot.png",
       plot = pH_plot,
       dpi = 1000,
       width = 20,
       height=12,
       units="cm")



##-----------Plotting the relationship of soil EC, soil depth and habitat type----------##

EC_post_hoc <- data.frame(
  Habitat_type = rep(c("Peatland", "Forest", "Grassland"), 2),
  Soil_depth = rep(c("0-15 cm","15-30 cm" ), each = 3),
  y = c(rep(400, 3),
        rep(500, 3)),
  Label = c("a,A", "a,A", "a,A",
            "a,A","a,A","a,A"))

EC_plot<-ggplot(soil_health_fixed, 
       aes(x = Habitat_type, y= EC, fill= Habitat_type)) + 
  xlab("")+
  ylab("Electrical conductivity μS/cm")+
  theme_bw()+
  facet_wrap(~Soil_depth)+
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))+
  geom_boxplot()+
  scale_fill_manual(values = c("#365D8DFF","#20A486FF","red"))+
  geom_text(data=EC_post_hoc,
            aes(y=y,
                x=Habitat_type,
                label=Label),
            color="black")+
  ggtitle("Soil electrical conductivity across soil depths and habitat types in Noorder park")+
    theme(plot.title = element_text(size = 10))

ggsave(filename = "EC_plot.png",
       plot = EC_plot,
       dpi = 1000,
       width = 20,
       height=12,
       units="cm")


```

## Physical characteristics

### fitting a GLM to physical characteristics

```{r}

#--------------------------------------------Bulk density---------------------------------#
soil_health_fixed <- soil_health_fixed|> 
  rename(Bulk_density = `Bulk density`)

Bulk_glm<-glm(Bulk_density~Habitat_type*Soil_depth, 
            data = soil_health_fixed, 
            family = Gamma(link = "log"))

Bulk_EMM<-emmeans(Bulk_glm,~Habitat_type*Soil_depth)

summary(Bulk_EMM)
pairs(Bulk_EMM)

null_deviance <- summary(EC_glm)$null.deviance
residual_deviance <- summary(EC_glm)$deviance

100*(null_deviance-residual_deviance)/null_deviance


#--------------------------------------------Soil moisture---------------------------------#


soil_health_fixed <- soil_health_fixed|> 
  rename(Soil_moisture= `Soil moisture`)

moist_glm<-glm(Soil_moisture~Habitat_type*Soil_depth, 
            data = soil_health_fixed, 
            family = Gamma(link = "log"))

moist_EMM<-emmeans(moist_glm,~Habitat_type*Soil_depth)

summary(moist_EMM)
pairs(moist_EMM)



null_deviance <- summary(EC_glm)$null.deviance
residual_deviance <- summary(EC_glm)$deviance

100*(null_deviance-residual_deviance)/null_deviance



#---------------------------------------Hydrophobicity----------------------------------------#

soil_health_fixed <- soil_health_fixed|> 
  rename(Hydrophobicity= `Water repellency  time(secs`) 

hydro_glm<-glm(Hydrophobicity~Habitat_type*Soil_depth, 
            data = soil_health_fixed, 
            family = Gamma(link = "log"))

hydro_EMM<-emmeans(hydro_glm,~Habitat_type*Soil_depth)

summary(hydro_EMM)
pairs(hydro_EMM)

null_deviance <- summary(EC_glm)$null.deviance
residual_deviance <- summary(EC_glm)$deviance

100*(null_deviance-residual_deviance)/null_deviance
```

------------------------------------------------------------------------

### Plotting physical characteristics

```{r}
#-----------------------------------------Soil bulk density---------------------------------#
bulk_post_hoc <- data.frame(
  Habitat_type = rep(c("Peatland", "Forest", "Grassland"), 2),
  Soil_depth = rep(c("0-15 cm","15-30 cm"), each = 3),
  y = c(rep(8, 3),
        rep(8, 3)),
  Label = c("a,A", "a,A", "a,A",
            "a,A","a,A","a,A"))


bulk_plot<-ggplot(soil_health_fixed, 
       aes(x = Habitat_type, y= Bulk_density, fill= Habitat_type)) + 
  xlab("")+
  ylab("Soil Bulk density (g/cm^3)")+
  theme_bw()+
  facet_wrap(~Soil_depth)+
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))+
  geom_boxplot()+
  scale_fill_manual(values = c("#365D8DFF","#20A486FF","red"))+
  geom_text(data=bulk_post_hoc,
            aes(y=y,
                x=Habitat_type,
                label=Label),
            color="black")+
  ggtitle("Soil bulk density across soil depths and habitat type in Noorder park")+
    theme(plot.title = element_text(size = 10))

ggsave(filename = "bulk_plot.png",
       plot = bulk_plot,
       dpi = 1000,
       width = 20,
       height=12,
       units="cm")

  

levels(soil_health_fixed$Site)
#---------------------------------------------Soil moisture---------------------------------#

moist_post_hoc <- data.frame(
  Habitat_type = rep(c("Peatland", "Forest", "Grassland"), 2),
  Soil_depth = rep(c("0-15 cm","15-30 cm" ), each = 3),
  y = c(rep(40, 3),
        rep(46, 3)),
  Label = c("a,A", "a,A", "a,A",
            "a,A","a,A","a,A"))


moist_plot<-ggplot(soil_health_fixed, 
       aes(x =Habitat_type, y=Soil_moisture, fill=Habitat_type)) + 
  xlab("")+
  ylab("Soil moisture content (%)")+
  theme_bw()+
  facet_wrap(~Soil_depth)+
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))+
  geom_boxplot()+
  scale_fill_manual(values = c("#365D8DFF","#20A486FF","red"))+
  geom_text(data=moist_post_hoc,
            aes(y=y,
                x=Habitat_type,
                label=Label),
            color="black")+
  ggtitle("Soil moisture content across soil depths and habitat types in Noorder park")+
    theme(plot.title = element_text(size = 10))



ggsave(filename = "moist_plot.png",
       plot = moist_plot,
       dpi = 1000,
       width = 20,
       height=12,
       units="cm")


#---------------------------------------------Hydrophobicity---------------------------------#
hydro_post_hoc <- data.frame(
  Habitat_type = rep(c("Peatland", "Forest", "Grassland"), 2),
  Soil_depth = rep(c("0-15 cm","15-30 cm" ), each = 3),
  y = c(rep(610, 3),
        rep(300, 3)),
  Label = c("a,A", "b,A", "a,A",
            "a,A","b,A","a,A"))


hydro_plot<-ggplot(soil_health_fixed, 
       aes(x = Habitat_type, y= Hydrophobicity, fill= Habitat_type)) + 
  xlab("")+
  ylab("Water repellency time (secs)")+
  theme_bw()+
  facet_wrap(~Soil_depth)+
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))+
  geom_boxplot()+
  scale_fill_manual(values = c("#365D8DFF","#20A486FF","red"))+
  geom_text(data=hydro_post_hoc,
            aes(y=y,
                x=Habitat_type,
                label=Label),
            color="black")+
  ggtitle("Soil hydrophobicity across soil depths and habitat type in Noorder park")+
  theme(plot.title = element_text(size = 10))

ggsave(filename = "hydro_plot.png",
       plot = hydro_plot,
       dpi = 1000,
       width = 20,
       height=12,
       units="cm")



```

# Correlation Matrices

```{r}
#--------------------------------------------Peatland-----------------------------------------#
data_p<-peat_cor_data
peat_mat<-cor(data_p, use = "pairwise.complete.obs")
cor_peat_mat<-corr.test(data_p)$p


 peat<-corrplot(peat_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 0.5,
         p.mat = cor_peat_mat,
         insig="pch",
         col = c("brown","lightblue"),
         na.label="square",
         na.label.col = "white")
 
 
peat<-ggcorrplot(peat_mat,
            p.mat=cor_peat_mat,
            type = "lower",
            insig = "pch",
            lab=TRUE)

#--------------------------------------------Forest------------------------------------------#
data_f<-forest_cor_data
forest_mat<-cor(data_f)
cor_forest_mat<-corr.test(data_f)$p


forest<-corrplot(forest_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 0.5,
         p.mat = cor_forest_mat,
         insig="pch",
         col = c("brown","lightblue"))


forest<-ggcorrplot(forest_mat,
            p.mat=cor_forest_mat,
            type = "lower",
            insig = "pch",
            lab=TRUE)

#----------------------------------------Grassland-------------------------------------------#
data_g<-grass_cor_data
grass_mat<-cor(data_g)
cor_grass_mat<-corr.test(data_g)$p


grass<-corrplot(grass_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 0.5,
         p.mat = cor_grass_mat,
         insig="pch",
         col = c("brown","lightblue"),
         na.label = "square")

grass<-ggcorrplot(grass_mat,
            p.mat=cor_grass_mat,
            type = "lower",
            insig = "pch",
            lab=TRUE)
#----------------------------------------------All data-------------------------------------#
data_all<-all_cor_data
all_mat<-cor(data_all)
cor_all_mat<-corr.test(data_all)$p


all<-corrplot(all_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 0.5,
         p.mat = cor_all_mat,
         insig="pch",
         col = c("brown","lightblue"))

all<-ggcorrplot(all_mat,
            p.mat=cor_all_mat,
            type = "lower",
            insig = "pch",
            lab=TRUE)
#--------------------------------------Putting plots together---------------------------------#
pdf("final.pdf", width = 18, height = 18)
par(mfrow=c(2,2), oma = c(0, 0, 3, 0))
corrplot(all_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 1.5,
         p.mat = cor_all_mat,
         insig="pch",
         col = c("brown","lightblue"),
         main= "A", cex.main =2, mar = c(0, 0, 2, 0))
corrplot(peat_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 1.5,
         p.mat = cor_peat_mat,
         insig="pch",
         col = c("brown","lightblue"),
         na.label="square",
         na.label.col = "white",
         main = "B", cex.main =2, mar = c(0, 0, 2, 0))
corrplot(forest_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 1.5,
         p.mat = cor_forest_mat,
         insig="pch",
         col = c("brown","lightblue"),
         main = "C", cex.main =2, mar = c(0, 0, 2, 0))
corrplot(grass_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 1.5,
         p.mat = cor_grass_mat,
         insig="pch",
         col = c("brown","lightblue"),
         main = "D", cex.main =2, mar = c(0, 0, 2, 0))
mtext("Correlation matrices of soil health and species richnes for each habitat indivdiually and combined ", outer = TRUE, cex = 2, line = 0.5)



pdf("final.pdf", width = 16, height = 14)
grid <-matrix(c(1,1,1,
                2,3,4), nrow = 2, ncol =3, byrow=TRUE)
layout(grid)
corrplot(all_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 1.5,
         p.mat = cor_all_mat,
         insig="pch",
         col = c("brown","lightblue"),
         main = "A",  cex.main =2, mar = c(0, 0, 2, 0))
corrplot(peat_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 1.5,
         p.mat = cor_peat_mat,
         insig="pch",
         col = c("brown","lightblue"),
         na.label="square",
         na.label.col = "white",
          main = "B", cex.main =2, mar = c(0, 0, 2, 0))
corrplot(forest_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 1.5,
         p.mat = cor_forest_mat,
         insig="pch",
         col = c("brown","lightblue"),
          main = "C", cex.main =2, mar = c(0, 0, 2, 0))
corrplot(grass_mat,
         order = "original",
         type = "lower",
         tl.col = "black",
         tl.cex = 1.5,
         p.mat = cor_grass_mat,
         insig="pch",
         col = c("brown","lightblue"),
          main = "D", cex.main =2, mar = c(0, 0, 2, 0))
mtext("Correlation matrices of the relationship soil health and species richnes for each habitat indivdiually and combined ", outer = TRUE, cex = 2, line = 0.5)
```
